pirouette: Determining the error BEAST2 makes in inferring a phylogeny

### 2019-03-04 article.tex
Improved 'Introduction' and 'Description' sections

\verb;pirouette; is written in the R programming language (\cite{R}).

The goal of \verb;pirouette; is to measure the inference error BEAST2
makes for a given reconstructed phylogeny, within the context of
a simulation study on a (possibly new) diversification model. Such
a study will typically construct phylogenies for different combinations of the diversification model's parameters, to assess under which scenarios the error made by BEAST2 cannot be neglected. 
This usually requires many replicates.

To explain how \verb;pirouette; measures the inference error
BEAST2 makes, we first give two definitions of what
we mean by a generative model and by an inference model: 
a generative model is a combination of site model, clock model and tree prior,
used to simulate DNA alignments. An inference model is a combination of 
site model, clock model, tree prior, an optional node calibration prior
and setup of the Markov chain Monte Carlo (MCMC) algorithm, 
used to produce a posterior distribution of trees and model parameters. Node calibration priors, called 'most
recent common ancestor' (MRCA) priors within BEAST2, allow to create a time-calibrated phylogeny, by specifying a normal distribution around an expected time when the ancestors of two or more taxa diversified.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a true phylogeny (1a), 
    that is converted to an alignment (2a).
    \rampal{this is unclear; the reader does not know what an experiment is, why there are more, what filtering means, etc}
    The use supplies one or more experiments, which are a combination of 
    an inference model and the conditions in which it is actually used in
    inference.
    It is determined which experiments will be actually run (3a), 
    possibly based on the
    alignment (2a).
    The experiment that are actually performed, use the 
    alignment (2a) to create a BEAST2 posterior of phylogenies (4a),
    which is compared to the true phylogeny (1a) 
    using the nLTT statistic [\cite{janzen2015approximate}]. (5a) shows the nLTT of the true tree in red,
    the nLTT of all individual posterior trees in grey, and the average
    nLTT statistic of all posterior trees in black.
    Taking the differences between the true tree's nLTT and each posterior's
    nLTT results in an error distribution, here displayed as a histogram
    of error values (6a).
    Optionally, a twin phylogeny (1b) can be generated from a true
    phylogeny (1a), which then results in an error distribution (6b) with
    similar intermediate steps 
  }
  \label{fig:pipeline}
\end{figure}

@@

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a).
One can specify a DNA sequence
of any length at the root of the phylogeny, a DNA mutation rate, a
site (i.e. nucleotide substitution) model, 
a clock model, a random number generator (RNG) seed and a location
where the alignment is saved to. This step is relatively fast, but longer
DNA alignments will noticeably slow down the inference step.

In the second step selects (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a), the user can specify whether to use the generative model as the inference model or to select the best model from a set of inference models. 
When selecting the generative model, the site and clock model used in the alignment simulation are used in the inference. Furthermore, because the given phylogeny (on which the alignment is based)
may have followed any tree prior (i.e. speciation model), the user needs
to specify which tree prior is used in the inference. 
When selecting for the best
model, the alignment is used to find the inference model that has the
highest evidence (i.e. marginal likelihood) from a set of candidate inference models.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; [\cite{mcbette}], that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac,
therefore inference model selection is not available on Windows. Note that
there is the option to do model selection from the browser, 
using \verb;mcbette;. The computation for this model selection is 
performed on the servers of the Travis CI continuous integration service,
that restricts computation time to one hour for free users.

The third step infers the actual Bayesian posterior from the simulated 
alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a $\rightarrow$ 4a),
using the inference model(s) selected in the previous step. The user
can specify the additional parameters needed for the BEAST2 run, which
are the Markov-Chain Monte Carlo (MCMC) setup, 
an optional MRCA prior and an RNG seed \giovanni{This has been already said when defining the inference model. I suggest to define it only once in detail (like here) and just mention it the first time (something like: "An inference model is a combination of site model, clock model, tree prior as well as additional BEAST2 settings." in the description section, around line 98), which will probably look a bit more neat, as it's a bit clearer the parallel between alignment model and inference model.}.
The MCMC setup determines the number of posterior trees sampled.
An MRCA prior allows the inferred phylogeny to have a dated crown age.

@@

An optional step is to generate a 'twin tree' (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
The twin tree has the same topology as the given phylogeny, 
yet its branching times are simulated according to 
the tree prior used by BEAST2 to produce a posterior 
\richel{Should be switched to BD, pirouette Issue 161}.
The goal of this procedure is to generate a tree 
to use as a control for the main experiment.
We produce the twin tree according to the following procedure.

From the given (true) phylogeny, we maximize a birth-death 
likelihood [\cite{nee1994reconstructed}] to infer the standard 
birth-death rates, namely speciation rate $\lambda$ and extinction rate $\mu$.
We use $\lambda$ and $\mu$ to simulate a new birth-death tree, 
conditioned on having the same number of tips as the original tree, we store its branching times, and then  combine the branching times with the original tree's 
topology to obtain the twin tree.

@@

We start with a Yule (pure-birth) tree with five taxa and a crown age of ten.

@@

\iffalse
In the second step we state our experiments. In this context, we
define an experiment as a combination of an inference model
and the conditions under which a Bayesian inference is executed.
Within this example, we specify that our experiment uses the
generative model, will always be run and we ignore the evidence for our model,
for the inference model that is also the generative model (using a
JC site model, strict clock model and Yule tree prior).
\fi

In the second step we state our experiments. In this context, we
define an experiment as a combination of an inference model
and the conditions under which a Bayesian inference is executed.
Within this example, we specify that our experiment uses the
generative model (which means that the inference model will be a combination of JC site model, strict clock model and Yule tree prior), will always be run and we don't need to measure the evidence for our model.

@@

\verb;pirouette; has a nice plotting function:

>>>>

\begin{table}
  \begin{tabular}{ | l | l | }
    \hline
    \textbf{Term} & \textbf{Definition} \\ 
    \hline
    Clock model & Model for the DNA mutation rates \\
    Experiment & Combination of an inference model and the conditions to actually use it in inference \\
    Generative model & Combination of site model, clock model and tree prior, used to simulate a DNA alignment \\
    Inference model & Combination of site model, clock model, tree prior, MRCA prior and MCMC, used to infer a BEAST2 posterior \\
    MCMC & Setup of the Markov Chain Monte Carlo algorithm \\
    MRCA prior & Prior knowledge of a most recent common ancestor \\
    Site model & Model for the DNA nucleotide substitions \\
    Tree prior & Model for the macroevolutionary speciation process \\
    \hline
  \end{tabular}
  \caption{
    Definitions of the terms used in this article
  }
  \label{tab:definitions}
\end{table}

\verb;pirouette; is written in the R programming language (\cite{R}).
The goal of \verb;pirouette; is to measure the inference error BEAST2
makes for a given reconstructed phylogeny, within the context of
a simulation study on a (possibly new) diversification model. Such
a study will typically construct phylogenies for different 
combinations of the diversification model's parameters, 
to assess under which scenarios the error made by BEAST2 cannot be neglected.
Although a well-conducted theoretical study uses many replicates,
this article only shows examples with one replicate, 
for the sake of readability.

\richel{
  I do think this is a clumsy paragraph. I hope you can merge it
  in the text
}
To explain how \verb;pirouette; measures the inference error
BEAST2 makes, we first give two 
definitions (see table \ref{tab:definitions} for an overview of definitions):
(1) a generative model is a combination of a known tree prior (that generated
a given phylogeny) and a site and clock model (that generate an alignment 
from the given tree), (2) an inference model is a combination of 
site model, clock model, tree prior, an optional most recent 
common-ancestor (MRCA) prior (to create a time-calibrated phylogeny)
and a Markov chain Monte Carlo (MCMC) setup.
A generative model is a known/true model underlying a phylogeny 
and is used to simulate DNA alignments.
An inference model
is used to produce a posterior distribution of trees and model parameters.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{workflow.png}
  \caption{
    \texttt{pirouette} pipeline. 
    The pipeline starts from a true phylogeny (1a), 
    that is converted to an alignment (2a).
    The use supplies one or more experiments, in which an experiment is
    defined as a combination of 
    an inference model and the conditions in which it is actually used in
    inference.
    Based on the experiments and possibly the alignment (2a),
    experiments are selected to be actually used used in the next step (3a).
    Those experiments that are selected, use the   
    alignment (2a) and their inference model 
    to create a BEAST2 posterior of (parameter estimates and) phylogenies (4a).
    The posterior trees are compared to the true phylogeny (1a) 
    using the nLTT statistic [\cite{janzen2015approximate}]. 
    (5a) shows the nLTT of the true tree in red,
    the nLTT of all individual posterior trees in grey, and the average
    nLTT statistic of all posterior trees in black.
    Taking the differences between the true tree's nLTT and each posterior's
    nLTT results in an error distribution, here displayed as a histogram
    of error values (6a).
    Optionally, a twin phylogeny (1b) can be generated from a true
    phylogeny (1a), which then results in an error distribution (6b) with
    similar intermediate steps 
  }
  \label{fig:pipeline}
\end{figure}

@@

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a).
One can specify, most importantly, a root DNA sequence, 
a DNA mutation rate, a site model, and a clock model
\richel{
  as the DNA mutation rate is the same over all branches, 
  only the strict clock model makes sense. I suggest to
  remove the option to specify a clock model
}.
This step is relatively fast, but longer
DNA alignments will noticeably slow down the inference step.

The second step (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a)
starts from experiments as set up by the user. 
We define an experiment as the combination of an inference model 
and the conditions to actually use it in inference.
There a two types of inference models, which are
generative and candidate inference models.
An experiment uses a generative model if the inference model is known.
An experiment uses a candidate model if either the inference model is unknown,
or there is an interest in seeing how well non-generative models compete with
the generative model.
Typically, a generative model and/or a best candidate model are used 
in the actual inference.
When using an experiment with a generative model,
it must use the same site and clock model used in the alignment simulation,
as well as the tree prior that underlied the creation of the given phylogeny. 
When using experiments with candidate models,
\verb;pirouette; selects the best candidate based
on the evidence (i.e. marginal likelihood) for the model given the alignment.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;mcbette; [\cite{mcbette}], that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac,
therefore using the selection of candidate models is not available on Windows. 
Note that there is the option to do just the model selection on Windows 
from the browser using \verb;mcbette;, yet the computation for this model selection is
restricted to one hour for free users of the host (Travis CI) continuous
integration service.

The third step infers the actual Bayesian posteriors from the simulated 
alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 3a $\rightarrow$ 4a),
using the inference models from the experiments 
selected in the previous step. Per selected experiment, 
one BEAST2 posterior is inferred, 
using the \verb;babette; [\cite{bilderbeek2018babette}] R package.

@@

An optional step is to generate a 'twin tree' 
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
The twin tree has the same topology as the given phylogeny, 
yet its branching times are simulated according to 
a birth-death tree model.
\richel{Should be switched to BD, pirouette Issue 161}.
The goal of this procedure is to generate a tree
with a tree prior present in BEAST2, 
that is used as a control for the main experiment.
To produce such a birth-death tree, we maximize a birth-death 
likelihood [\cite{nee1994reconstructed}], 
conditioned on having the same number of tips as the original tree,
to infer its parameters.
We use these parameters,  
which are a speciation rate and extinction rate,
to simulate the branching times for the 
twin tree, which we combinate with the original tree's 
topology.

@@

\richel{Suggest: use BD tree (again)}
We start with a Yule (pure-birth) tree with five taxa and a crown age of ten.

@@

In the second step we state our experiments. In this context, we
define an experiment as a combination of an inference model
and the conditions under which a Bayesian inference is executed.
Within this example, we specify that our experiment uses the
generative model (which means that the inference model will be 
a combination of JC site model, strict clock model and Yule tree prior), 
will always be run and we skip to measure the evidence for our model.

@@

\verb;pirouette; has a plotting function used for convenience:
